Regression Diagnostics
An excellent review of regression diagnostics is provided in John Fox's aptly named Overview of Regression Diagnostics. Dr. Fox's car package provides advanced utilities for regression modeling.


# Assume that we are fitting a multiple linear regression
# on the MTCARS data
library(car)
fit <- lm(mpg~disp+hp+wt+drat, data=mtcars)
This example is for exposition only. We will ignore the fact that this may not be a great way of modeling the this particular set of data!

\subsection*{Outliers}

# Assessing Outliers
outlierTest(fit) # Bonferonni p-value for most extreme obs
qqPlot(fit, main="QQ Plot") #qq plot for studentized resid 
leveragePlots(fit) # leverage plots
leverage plot click to view


\subsection*{Influential Observations}
# Influential Observations
# added variable plots 
av.Plots(fit)

# Cook's D plot
# identify D values > 4/(n-k-1) 
cutoff <- 4/((nrow(mtcars)-length(fit$coefficients)-2)) 
plot(fit, which=4, cook.levels=cutoff)

# Influence Plot 
influencePlot(fit,	id.method="identify", main="Influence Plot", sub="Circle size is proportial to Cook's Distance" )
av plots Cook's D plot influence plot click to view

\subsection*{Non-normality}
# Normality of Residuals
# qq plot for studentized resid
qqPlot(fit, main="QQ Plot")
# distribution of studentized residuals
library(MASS)
sresid <- studres(fit) 
hist(sresid, freq=FALSE, 
   main="Distribution of Studentized Residuals")
xfit<-seq(min(sresid),max(sresid),length=40) 
yfit<-dnorm(xfit) 
lines(xfit, yfit)
qq plot histogram of studentized residuals click to view

\subsection{Non-constant Error Variance}
# Evaluate homoscedasticity
# non-constant error variance test
ncvTest(fit)
# plot studentized residuals vs. fitted values 
spreadLevelPlot(fit)
spread vs. levels click to view
\subsection*{Multi-collinearity}
# Evaluate Collinearity
vif(fit) # variance inflation factors 
sqrt(vif(fit)) > 2 # problem?

\subsection*{Nonlinearity}
# Evaluate Nonlinearity
# component + residual plot 
crPlots(fit)
# Ceres plots 
ceresPlots(fit)
component plus residual plot Ceres plots click to view
Non-independence of Errors
# Test for Autocorrelated Errors
durbinWatsonTest(fit)

Additional Diagnostic Help
The gvlma( ) function in the gvlma package, performs a global validation of linear model assumptions as well separate evaluations of skewness, kurtosis, and heteroscedasticity.
# Global test of model assumptions
library(gvlma)
gvmodel <- gvlma(fit) 
summary(gvmodel)
