Central Limit Theorem
========================================================



Hypothesis testing and confidence interval construction are based on the ***Central Limit Theorem*** (CLT).

Recall the Dice experiment.
```{r}
Dice1=floor(runif(50,min=1,max=7))  
```
A simple demonstration of the CLT is given by the problem of rolling a large number of dice repeatedly. 
The distribution of the sum (or average) of the rolled numbers will be well approximated by a normal distribution, the parameters of which can be determined empirically.
```{r}
N=1000            #number of loops
Avgs=numeric(N)  #array “Avgs” store the sample means
for( i in 1:N)
     { 
     Dice=floor(runif(50,min=1,max=7));
     Avgs[i]=mean(Dice)  
}                                 
Avgs[1:20]            #print first 20 values from Avgs dataset to screen
``` 

Lets look at the distribution of the means. Are they normally distributed?
```{r}
mean(Avgs)          #compute the mean. 
qqnorm(Avgs,pch=16,col="red")        #draws a QQ plot
qqline(Avgs,col="blue")        #adds trend line to QQplot.
shapiro.test(Avgs)  #Shapiro Wilk test. 
```
```{r}
hist(Avgs,breaks=20,col="lightblue")
abline(v=3.5,col="red")
```


